---
title: "DataCollection"
author: "Liz Conley"
date: "2026-04-15"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readxl)
library(lubridate)
library(tidycensus)
library(sf)
library(dplyr)
```

#Shapefiles for metro stations & lines from from [ArcGIS Hub](https://hub.arcgis.com/maps/mwcog::metro-lines-and-stations-1/about?path=):
Code block hidden since I've cached this all to a csv.
//```{r}
stations <- st_read("RawData/MetroShapefiles/stations/Metro_Rail_Stations.shp") #station names cleaned below
lines <- st_read("RawData/MetroShapefiles/lines/Metro_Rail_Lines.shp") #station names not cleaned
///```

List of cleaned station names:
Code block hidden since I've cached this all to a csv.
//```{r}
stations[96,3] <- "College Park-U of Md"
stations[96,4] <- "College Park-U of Md"

stations[66,3] <- "U StreetAfrican-Amer Civil War MemorialCardozo"
stations[66,4] <- "U StreetAfrican-Amer Civil War MemorialCardozo"

stations[50,3] <- "Woodley Park-ZooAdams Morgan"
stations[50,4] <- "Woodley Park-ZooAdams Morgan"

stations[33,3] <- "ViennaFairfax-GMU"
stations[33,4] <- "ViennaFairfax-GMU"


StationNames <- stations %>% select(NAME) %>% arrange(NAME)

//```

#Adjacency Matrix for Stations
```{r}
#adj <- as.matrix(read_csv("RawData/railme/adj.csv", col_names=FALSE)) #adjacency matrix for stations
#matrix number not the same as station number from shapefile, correct mapping is in nodes.

#nodes <- read_csv("RawData/railme/nodes.csv", col_names=FALSE) %>% arrange(X2) %>% rename(MatrixNumber = X1) %>% rename(NAME = X2)
#contains matrix number mappings for the stations.

#stations2 <- stations %>% arrange(NAME)
#stations3 <- full_join(nodes, stations2) %>% arrange(MatrixNumber)
#stations3 <- st_as_sf(stations3, coords=c("XCoord", "YCoord"))
#Shapefile with correct matrix numbers.

#stations3 <- stations3 %>% rename("Station" = "NAME")

#st_write(stations3, "CleanData/StationsShapefile.shp", append=FALSE)

StationsAdjacencyMatrix <- as.matrix(read_csv("CleanData/StationsAdjacencyMatrix.csv", col_names=FALSE))

CleanedStations <- st_read("CleanData/StationsShapefile.shp")
```


#Population Data
```{r}

```


#[Daily Rail Entries over Time](https://www.wmata.com/initiatives/ridership-portal/Metrorail-Ridership-Summary.cfm):
Year: 2023-2025
Date: 1/1/2023 - 03/31/2026
Month: all (Jan-Dec)
Day of Week: all (Sun-Sat)
Service Type: all (Saturday, Sunday, Weekday, July 4th Special, Sunday Reduced)
Time Period: all (AM Peak - Late Night)
Holiday: all (no & yes)
Code block hidden since I've cached this all to a csv.
//r}
read_DEwS <- function(StationName){
  path <- str_c("RawData/Daily Rail Entries over Time/",StationName,".xlsx", sep="")
  data <- read_excel(path) %>% 
    pivot_longer(cols=!`Service Type`, names_to="Date", values_to="Ridership") %>%
    na.omit() %>%
    mutate(Station=StationName) %>%
    select(Ridership, Date, Station, `Service Type`) %>%
    rename(Service = `Service Type`) %>%
    mutate(Date = as.Date(mdy(Date)))
}

read_all_DEwS <- function(StationNames){
  DEwS <- read_DEwS(StationNames[1])
  for(x in 2:98){
    temp_DEwS <- read_DEwS(StationNames[x])
    DEwS <- rbind(DEwS, temp_DEwS)
  }
  DEwS
}

DEwS <- read_all_DEwS(StationNames$NAME)
//```

#[Total Entries and Exits by Day](https://www.wmata.com/initiatives/ridership-portal/Metrorail-Ridership-Summary.cfm):
Year: 2023-2025
Date: 1/1/2023 - 03/31/2026
Month: all (Jan-Dec)
Day of Week: all (Sun-Sat)
Service Type: all (Saturday, Sunday, Weekday, July 4th Special, Sunday Reduced)
Time Period: all (AM Peak - Late Night)
Holiday: all (no & yes)
Code block hidden since I've cached this all to a csv.
//```{r}
read_DEaX <- function(StationName){
  path <- str_c("RawData/Total Entries and Exits by Day/",StationName,".xlsx", sep="")
  data <- read_excel(path) %>% 
  mutate(Station=StationName) %>%
  rename(Date = "...1") %>% mutate(Date = as.Date(mdy(Date))) %>%
  select(Entries, Exits, Date, Station)
}

read_all_DEaX <- function(StationNames){
  DEaX <- read_DEaX(StationNames[1])
  for(x in 2:98){
    temp_DEaX <- read_DEaX(StationNames[x])
    DEaX <- rbind(DEaX, temp_DEaX)
  }
  DEaX
}

DEaX <- read_all_DEaX(StationNames$NAME)
//```

Daily Entries and Exits with Service:
Code block hidden since I've cached this all to a csv.
//```{r}
DEaXwS <- merge(DEaX, DEwS) %>% select(!Ridership)

write.csv(DEaXwS, "RawData/DEaXwS.csv")

rm(DEaX, DEwS)
//```
```{r}
DEaXwS <- read_csv("RawData/DEaXwS.csv") %>% select(c(Date, Station, Entries, Exits, Service))
```

#East Coast gas prices (weekly):
```{r}
GasPrices <- read_excel("RawData/GasPrices/1b.xls", sheet=2, range="A3:C1863") %>%
  mutate(Date = as.Date(ymd(Date))) %>%
  filter(Date >= "2023-01-01") %>% filter(Date < "2026-03-31") %>%
  rename(`East Coast Gas Price` = `Weekly East Coast Regular Conventional Retail Gasoline Prices  (Dollars per Gallon)`) %>%
  select(Date, `East Coast Gas Price`) %>%
  add_row(Date=as.Date("2023-01-01"), `East Coast Gas Price`=3.008, .before=1) %>%
  add_row(Date=as.Date("2023-01-03"), `East Coast Gas Price`=3.172, .before=3) %>%
  add_row(Date=as.Date("2023-01-04"), `East Coast Gas Price`=3.172, .before=4) %>%
  add_row(Date=as.Date("2023-01-05"), `East Coast Gas Price`=3.172, .before=5) %>%
  add_row(Date=as.Date("2023-01-06"), `East Coast Gas Price`=3.172, .before=6) %>%
  add_row(Date=as.Date("2023-01-07"), `East Coast Gas Price`=3.172, .before=7) %>%
  add_row(Date=as.Date("2023-01-08"), `East Coast Gas Price`=3.172, .before=8) %>%
  complete(Date = full_seq(Date, 1))%>% 
  fill(`East Coast Gas Price`) %>%
  add_row(Date=as.Date("2026-03-31"), `East Coast Gas Price`=3.789)
```

#Dates of gov't shutdown:
```{r}
GovtShutdown <- read_csv("RawData/CityEvents/GovtShutdown.csv")
```

#Dates for cherry blossoms festival, from [National Cherry Blossom Festival](https://nationalcherryblossomfestival.org/):
```{r}
CherryBlossomsFestival <- read_csv("RawData/CityEvents/CherryBlossomsFestival.csv") %>%
  mutate(Date = mdy(Date))

```

#Dates for caps games, from [HockeyReference](https://www.hockey-reference.com/teams/WSH/2024_games.html):
```{r}
hockey <- read_excel("RawData/CityEvents/CapsGames/Caps.xlsx")
hockey$Location <- replace_na(hockey$Location, "home")
hockey <- hockey %>% filter(Location=="home") %>% select(-GP)
```

#All time series data combined into 1 data set
```{r}
TimeSeries <- merge(GasPrices, GovtShutdown)
TimeSeries <- merge(TimeSeries, CherryBlossomsFestival)
TimeSeries2 <- full_join(TimeSeries, hockey, by="Date") %>% filter(Date<=as.Date("2026-03-31"))
TimeSeries2$Location <- replace_na(TimeSeries2$Location, "none")
TimeSeries2$Season <- replace_na(TimeSeries2$Season, "none")
TimeSeries2 <- TimeSeries2 %>% rename("HockeyLocation" = "Location")
TimeSeries2 <- TimeSeries2 %>% rename("HockeySeason" = "Season")
```

#All combined into 1 data set
```{r}
DataSet <- full_join(DEaXwS, TimeSeries2, by="Date")
write_csv(DataSet, file="CleanData/DataSet.csv", append=FALSE, col_names=TRUE)

DataSet <- read_csv("CleanData/DataSet.csv")

#may want to morph data from daily to average weekly, flatten out noise & also reduce computational load?
#https://www.statology.org/aggregate-daily-data-in-r/
library(data.table)
DataSet_WeeklyAvg <- DataSet

#Encode special events as numeric, for a weekly score out of 7 for how many days of the week the event was happening
DataSet_WeeklyAvg$Shutdown <- as.numeric(DataSet_WeeklyAvg$Shutdown)
DataSet_WeeklyAvg$CherryBlossoms <- as.numeric(DataSet_WeeklyAvg$CherryBlossoms)

#Drop hockey season, games are already rare enough that I don't think the season will have a strong effect

library(dplyr)
DataSet_WeeklyAvg$HockeyLocation <- DataSet_WeeklyAvg$HockeyLocation %>% 
  mutate_if(is.character, as.factor)


dt <- as.data.table(DataSet_WeeklyAvg)
Entries_WeeklyAvg <- dt[, .(weekly_mean = mean(Entries)), 
                 by = .(week = floor_date(Date, "week"))]
Exits_WeeklyAvg <- dt[, .(weekly_mean = mean(Entries)), 
                 by = .(week = floor_date(Date, "week"))]
GasPrice_WeeklyAvg <- dt[, .(weekly_mean = mean(`East Coast Gas Price`)), 
                 by = .(week = floor_date(Date, "week"))]
```


#NEED TO ENTER DATES
Dates for capital pride, from []():
```{r}
CapitalPride <- read_csv("RawData/CityEvents/CapitalPride.csv")
```

#NEED TO ENTER DATES
Metro planned construction, from [WMATA](https://www.wmata.com/initiatives/plans/):
```{r}

```

#NEED TO EXTRACT DATA
Service Excellence Report Data File from WMATA's [Public Records](https://www.wmata.com/about/records/public-records.cfm#performance):
```{r}

```

#NEED TO COLLECT/EXTRACT DATA
WMATA Pedshed & Bikeshed Atlas Scores from [Pedshed & Bikeshed Atlas](https://www.wmata.com/initiatives/pedshed-and-bikeshed-atlas/):
```{r}

```

#NEED TO COLLECT DATA
Weather from []():
```{r}

```
