---
title: "MetroData"
author: "Liz Conley"
date: "2026-04-09"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readxl)
library(lubridate)
library(tidycensus)
```

#Step 1: Data Collection & Cleaning

##Response Data
This data comes from [WMATA's MetroRail Ridership Summary](https://www.wmata.com/initiatives/ridership-portal/Metrorail-Ridership-Summary.cfm).
[This doccument](https://www.wmata.com/initiatives/ridership-portal/upload/RDP-Metadata.pdf) is helpful for understanding how the data comes from WMATA.

---

Elements used in following steps:
```{r}
# Fort Totten is in there twice, as B09 and B06
# Gallery Pl-Chinatown is in there twice, as B01 and F01
# L'Enfant Plaza is D03 & F03
# Metro Center is on there twice as A01 and A10


# The stations that are in there 2x are stations w/ multiple levels & connections to multiple lines


StationList <- read_csv("wmata-rail-stations.csv", show_col_types = FALSE) %>% 
  rename(Station = Name) %>%
  arrange(Station) %>%
  select(Station, Code)

StationNames <- unique(StationList[[1]])
#StationCodes <- StationList[[2]]

rm(StationList)
```


---
This dataset is good, but only has entrances, not exits. Also have to get a different .xlsx for each station.

Source: [Daily Rail Entries over Time](https://www.wmata.com/initiatives/ridership-portal/Metrorail-Ridership-Summary.cfm)
Year: 2023-2025
Date: 1/1/2023 - 03/31/2026
Month: all (Jan-Dec)
Day of Week: all (Sun-Sat)
Service Type: all (Saturday, Sunday, Weekday, July 4th Special, Sunday Reduced)
Time Period: all (AM Peak - Late Night)
Holiday: all (no & yes)
```{r}
#Daily Entries with Service _ StationName
DEwS_AddisonRoad <- read_excel("Daily Rail Entries over Time/Addison Road-Seat Pleasant.xlsx") %>% 
  pivot_longer(cols=!`Service Type`, names_to="Date", values_to="Ridership") %>%
  na.omit() %>%
  mutate(Station="Addison Road-Seat Pleasant") %>%
  select(Ridership, Date, Station, `Service Type`) %>%
  rename(Service = `Service Type`) %>%
  mutate(Date = as.Date(mdy(Date)))

rm(DEwS_AddisonRoad)
```
ridership: how many people entered the system
station: the station they entered the system through
date: the date at which they entered the system
service: the service that was being offered at that date (system wide)

Now as a function:
```{r}
read_DEwS <- function(StationName){
  path <- str_c("Daily Rail Entries over Time/",StationName,".xlsx", sep="")
  data <- read_excel(path) %>% 
    pivot_longer(cols=!`Service Type`, names_to="Date", values_to="Ridership") %>%
    na.omit() %>%
    mutate(Station=StationName) %>%
    select(Ridership, Date, Station, `Service Type`) %>%
    rename(Service = `Service Type`) %>%
    mutate(Date = as.Date(mdy(Date)))
}

read_all_DEwS <- function(StationNames){
  DEwS <- read_DEwS(StationNames[1])
  for(x in 2:98){
    temp_DEwS <- read_DEwS(StationNames[x])
    DEwS <- rbind(DEwS, temp_DEwS)
  }
  DEwS
}

DEwS <- read_all_DEwS(StationNames)
```

---

This dataset is better, has entrances and exits. Still have to get a different .xlsx for each station.
However, it doesn't record service type.

Source: [Total Entries and Exits by Day](https://www.wmata.com/initiatives/ridership-portal/Metrorail-Ridership-Summary.cfm)
Year: 2023-2025
Date: 1/1/2023 - 03/31/2026
Month: all (Jan-Dec)
Day of Week: all (Sun-Sat)
Service Type: all (Saturday, Sunday, Weekday, July 4th Special, Sunday Reduced)
Time Period: all (AM Peak - Late Night)
Holiday: all (no & yes)
```{r}
#Daily Entries and Exits _ StationName
DEaX_AddisonRoad <- read_excel("Total Entries and Exits by Day/AddisonRoad.xlsx") %>%
  mutate(Station="AddisonRoad") %>%
  mutate(Code="G03") %>%
  rename(Date = "...1") %>% mutate(Date = as.Date(mdy(Date))) %>%
  select(Entries, Exits, Date, Station, Code)
  
```

Now as a function:
```{r}
read_DEaX <- function(StationName){
  path <- str_c("Total Entries and Exits by Day/",StationName,".xlsx", sep="")
  data <- read_excel(path) %>% 
  mutate(Station=StationName) %>%
  rename(Date = "...1") %>% mutate(Date = as.Date(mdy(Date))) %>%
  select(Entries, Exits, Date, Station)
}

read_all_DEaX <- function(StationNames){
  DEaX <- read_DEaX(StationNames[1])
  for(x in 2:98){
    temp_DEaX <- read_DEaX(StationNames[x])
    DEaX <- rbind(DEaX, temp_DEaX)
  }
  DEaX
}

DEaX <- read_all_DEaX(StationNames)
```

---

Now I'm going to combine the 2 data sets:

```{r}
#Daily Entries and Exits with Service
DEaXwS <- merge(DEaX, DEwS) %>% select(!Ridership)

rm(DEaX, DEwS)
```



## Explanitory Data
According to [these slides](https://ocw.mit.edu/courses/1-258j-public-transportation-systems-spring-2017/30236c5948850df0b1e4785cabb83b7f_MIT1_258JS17_lec08.pdf), some of the biggest factors affecting ridership are:

1. Exogenous (uncontrollable)
a. auto ownership/availability & operating costs
b. fuel prices & availability
c. demographics (age, gender, etc.) (age: young ppl don't have drivers licenses, old ppl struggle to drive. race: racial disparities track onto wealth disparities, track onto car ownership vs public transportation use.)
d. activity system (population & employment distributions, etc.)

2. Endogenous (controllable)
a. fare
b. headway (wait time)
c. route structure (walk time; ride time)
d. crowding*
e. reliability*

Some other predictive factors I could think of are:

3. Other factors I thought of (mix of Exogenous and Endogenous)
a. weather
b. bus
c. traffic
d. bike/pedshed info
e. time series data (holidays, schools in session or not, gov't shutdown)
f. distance from center of network (metro center) (crow's miles, rail miles, stops, etc?)
g. planned construction

---

1a.
Data on auto ownership/availability & operating costs: (none acquired yet.) 
Could use the US Census Beauro Vehicle Inventory and Use Survey (VIUS), but the Annual granularity is likely too broad.

---

1b.
Data on fuel prices & availability: (none acquired yet.) 
Source: US EIA (Energy Information Administration) [U.S. Regular Gasoline Prices*(dollars per gallon)](https://www.eia.gov/petroleum/gasdiesel/)

```{r}
GasPrices <- read_excel("ExplData/1b.xls", sheet=2, range="A3:C1863") %>%
  mutate(Date = as.Date(ymd(Date))) %>%
  filter(Date >= "2023-01-01") %>% filter(Date < "2026-03-31") %>%
  rename(`East Coast Gas Price` = `Weekly East Coast Regular Conventional Retail Gasoline Prices  (Dollars per Gallon)`) %>%
  select(Date, `East Coast Gas Price`) %>%
  add_row(Date=as.Date("2023-01-01"), `East Coast Gas Price`=3.008, .before=1) %>%
  add_row(Date=as.Date("2023-01-03"), `East Coast Gas Price`=3.172, .before=3) %>%
  add_row(Date=as.Date("2023-01-04"), `East Coast Gas Price`=3.172, .before=4) %>%
  add_row(Date=as.Date("2023-01-05"), `East Coast Gas Price`=3.172, .before=5) %>%
  add_row(Date=as.Date("2023-01-06"), `East Coast Gas Price`=3.172, .before=6) %>%
  add_row(Date=as.Date("2023-01-07"), `East Coast Gas Price`=3.172, .before=7) %>%
  add_row(Date=as.Date("2023-01-08"), `East Coast Gas Price`=3.172, .before=8) %>%
  complete(Date = full_seq(Date, 1))%>% 
  fill(`East Coast Gas Price`) %>%
  add_row(Date=as.Date("2026-03-31"), `East Coast Gas Price`=3.789)
```

---

1c.
Data on demographics: (can use tidycensus for demographic data on VA, DC & MD.)
```{r}
#acs <- get_acs(geography = "county", variables = )
```

---

1d. activity system (population & employment distributions, etc.):
part of this (population) falls under demographics.
employment data could be gotten from ?

Dates of gov't shutdown could be useful:
```{r}
temp_GovtShutdown1 <- GasPrices %>%
  mutate(Shutdown = FALSE) %>%
  select(Date, Shutdown)

temp_GovtShutdown2 <- temp_GovtShutdown1 %>%
  filter(between(Date, as.Date('2025-10-01'), as.Date('2025-11-12'))) %>%
  mutate (Shutdown = TRUE)

temp_GovtShutdown3 <- temp_GovtShutdown1 %>%
  filter(between(Date, as.Date('2026-01-31'), as.Date('2026-02-03'))) %>%
  mutate (Shutdown = TRUE)

temp_GovtShutdown1 <- temp_GovtShutdown1 %>%
  filter(!between(Date, as.Date('2025-10-01'), as.Date('2025-11-12'))) %>%
  filter(!between(Date, as.Date('2026-01-31'), as.Date('2026-02-03')))

GovtShutdown <- full_join(temp_GovtShutdown1, temp_GovtShutdown2) %>% full_join(temp_GovtShutdown3) %>% arrange(Date)

rm(temp_GovtShutdown1, temp_GovtShutdown2, temp_GovtShutdown3)
```

Dates for cherry blossoms could also be useful, use the dates of the [National Cherry Blossom Festival](https://nationalcherryblossomfestival.org/).
2023: 2023-03-20 to 2023-04-16
2024: 2024-03-20 to 2024-04-14
2025: 2025-03-20 to 2025-04-13
2026: 2026-03-20 to 2026-04-12
```{r}

```

Dates for caps games could also be useful, can be gotten from [HockeyReference](https://www.hockey-reference.com/teams/WSH/2024_games.html).
```{r}
hockey <- read_excel("CapsGames/Caps.xlsx")

hockey$location <- replace_na(hockey$Location, "home")
```


Dates for capital pride could also be useful:

Dates for planned construction could also be useful:


---

2a. fare:
Metro fares are difficult to determine without knowing exact entrance and exit.
The default flat fare price could be used?
Service may encompass some of this, as fares vary with service.

Use Service from DEaWxS.

---

2b. headway (wait time)
Station and line specific.
Might be able to get this data from google maps?

See 2e. reliability*

---

2c. route structure (walk time; ride time)
Also difficult to determine without knowing exact route.
Could maybe factor station pedshed score into this?

---

2d. crowding*
Station and line specific.
Might be able to get this data from google maps?

Best source is prolly the Service Excellence Report Data File from WMATA's [Public Records](https://www.wmata.com/about/records/public-records.cfm#performance),
specifically the RCROW tab.

---

2e. reliability*
Station and line specific.
Might be able to get this data from google maps?
Could get list of planned & unplanned station construction? And/or planned bus detour dates?

Best source is prolly the Service Excellence Report Data File from WMATA's [Public Records](https://www.wmata.com/about/records/public-records.cfm#performance)
specifically RAOTP (rail on time performance), RSDEL (percentage of planned services delivered), ELAVA (elevator availability) and ESAVA (escelator availability).
Also look at REILA (fleet reliability)

---

3a. weather

---

3b. bus

---

3c. traffic

---

3d. bike/pedshed info

---

3e. time series data (holidays, schools in session or not, gov't shutdown)

---

3f. distance from center of network (metro center) (crow's miles, rail miles, stops, etc?)

---

3g. planned construction
See WMATA Initiaves -> [Planned Construction](https://www.wmata.com/initiatives/plans/)

---
